Synthesizing Quotient Lenses

### 2018-03-15 papers/QuotientLenses/main.tex
more polishing

Strong unambiguity is necessary in the premises of the regular QRE combinators
because it ensures that the canonizing function of a QRE is well-defined.
For example, consider the QRE \lstinline{"a"* . (collapse "a"$^*$ $\to$ "a")}. The
behaviour of this QRE is not well-defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"}
depending on how \cd{"aaa"} is parsed. We also require that the regular
combinators applied to the kernels of QREs be unambiguous since the underlying
bijective string lens of a QRE lens operates on kernels, and bijective string lenses
impose the same unambiguity restrictions so that they too are well defined as
functions.

@@

Recall that given regular expressions $R, S$ and equivalence relations
$\equiv_R$ and $\equiv_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$, a
{\em bijective quotient lens} $q : R /{\equiv_R} \Leftrightarrow S/{\equiv_S}$
is a pair of functions $q.\get : \mathcal{L}(R) \longrightarrow \mathcal{L}(S)$
and $q.\lput : \mathcal{L}(S) \longrightarrow \mathcal{L}(R)$ such that for all
$r \in \mathcal{L}(R)$ and $s \in \mathcal{L}(S)$, we have $q.\lput(q_1.\get(r))
\equiv_R r$ and $q.\get(q_1.\lput(s)) \equiv_R s$. Moreover, if $r \equiv_R
r'$, then $q.\get(r) = q.\get(r')$, and if $s \equiv_S s'$, then $q.\lput(s) =
q.\lput(s')$. In words, a bijective quotient lens $q$ from $R$ to $S$ modulo
$\equiv_R$ and $\equiv_S$ is a pair of functions $q.\get$ and $q.\lput$ such
that $q.\get$ respects $\equiv_R$ and $q.\lput$ respects $\equiv_S$, and such
that the {\em get} and {\em put} functions lifted to $\mathcal{L}(R)/{\equiv_R}$
and $\mathcal{L}(S)/{\equiv_S}$ are mutual inverses. These laws are similar
bijective lens laws \cref{bijectivelenslaws}, except that the equality
restrictions in the bijective lens laws are loosened to allow for equivalence
relations. Also, the condition that $r \equiv_R r'$ implies $q.\get(r) =
q.\get(r')$ ensures that the {\em get} function induced on the equivalence
classes if $\equiv_c$ is well-defined, and similarly for the {\em put}
function.

@@

We have implemented QREs and the quotient lens synthesis algorithm described
above as an extension to the Boomerang
interpreter~\cite{boomerang,quotientlenses}. We have extended the existing
Optician tool to synthesize QRE lenses from QREs. We will use word
\emph{\QOpt} to denote our extended version of Optician, and Optician to denote
the pre-QRE version of Optician. The synthesis algorithm produces Boomerang lens
values, so Boomerang treats synthesized lenses the same as hand-written ones. To
evaluate the effectiveness of QREs, QRE lenses, and QRE lens synthesis, we
conducted experiments to answer the following questions:
%
\begin{itemize}
  \item {\bf Ease of use.} Does synthesizing QRE lenses from QREs permit an
  easier development process than writing lenses by hand?
  Does synthesizing QRE lenses from QREs permit an easier development
  process than manually writing canonizers and then synthesizing lenses between
  their canonized forms?  
  \item {\bf Performance.} Is the synthesis algorithm/implementation fast enough to be
  used as part of a standard development process?
\end{itemize}
%
All evaluations were performed on a 2.5 GHz Intel Core i7 processor with 16 GB
of 1600 MHz DDR3 running macOS High Sierra.

@@

\BCP{Wrote the rest---comments?}\saz{This seems OK---maybe swap the two
  paragraphs.  The second one sounds more incremental and less exciting than the
  first, so it's maybe not as good of a note to end on?}
Looking ahead, we would like to investigate whether the techniques proposed
here can be applied to synthesizing quotiented variants of richer classes of
lenses, such as ``classic'' asymmetric lenses~\cite{FosterLenses} or
symmetric lenses~\cite{symmetriclenses}.  The main technical issues are
developing synthesis procedures for the unquotientied variants of a given
class---we are currently attacking the case of symmetric lenses---and
showing that quotiented lenses in this class can be normalized along the
lines of Theorem~\ref{normal form}.

Along a different dimension, we are experimenting with larger examples of
quotiented bijective lenses.  So far, the three ``specialized'' QRE
combinators (\kw{squash}, \kw{perm}, and \kw{collapse}) have proved
sufficient, but we expect eventually to encounter examples that will lead us
to consider adding further combinators to supplement these.

>>>>

Strong unambiguity is necessary in the premises of the regular QRE combinators
because it ensures that the canonizing function of a QRE is well-defined.
For example, consider the QRE \lstinline{"a"$^*$ . (collapse "a"$^*$ $\to$ "a")}. The
behavior of this QRE is not well-defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"}
depending on how \cd{"aaa"} is parsed. We also require that the regular
combinators applied to the kernels of QREs be unambiguous since the underlying
bijective string lens of a QRE lens operates on kernels, and bijective string lenses
impose the same unambiguity restrictions so that they too are well defined as
functions.

@@

Recall that given regular expressions $R, S$ and equivalence relations
$\equiv_R$ and $\equiv_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$, a
{\em bijective quotient lens} $q : R /{\equiv_R} \Leftrightarrow S/{\equiv_S}$
is a pair of functions $q.\get : \mathcal{L}(R) \longrightarrow \mathcal{L}(S)$
and $q.\lput : \mathcal{L}(S) \longrightarrow \mathcal{L}(R)$ such that for all
$r \in \mathcal{L}(R)$ and $s \in \mathcal{L}(S)$, we have $q.\lput(q_1.\get(r))
\equiv_R r$ and $q.\get(q_1.\lput(s)) \equiv_R s$. Moreover, if $r \equiv_R
r'$, then $q.\get(r) = q.\get(r')$, and if $s \equiv_S s'$, then $q.\lput(s) =
q.\lput(s')$. In words, a bijective quotient lens $q$ from $R$ to $S$ modulo
$\equiv_R$ and $\equiv_S$ is a pair of functions $q.\get$ and $q.\lput$ such
that $q.\get$ respects $\equiv_R$ and $q.\lput$ respects $\equiv_S$, and such
that the {\em get} and {\em put} functions lifted to $\mathcal{L}(R)/{\equiv_R}$
and $\mathcal{L}(S)/{\equiv_S}$ are mutual inverses. These laws are similar
bijective lens laws \cref{bijectivelenslaws}, except that the equality
restrictions in the bijective lens laws are loosened to allow for equivalence
relations. Also, the condition that $r \equiv_R r'$ implies $q.\get(r) =
q.\get(r')$ ensures that the \kw{get} function induced on the equivalence
classes of $\equiv_R$ is well-defined, and similarly for the \kw{put}
function.

@@

We have implemented QREs and the quotient lens synthesis algorithm described
above as an extension to the Boomerang
interpreter~\cite{boomerang,quotientlenses}. We have extended the existing
Optician tool to synthesize QRE lenses from QREs. We will use 
``\QOpt'' to denote our extended version of Optician, and just plain ``Optician'' to denote
the pre-QRE version of Optician. The synthesis algorithm produces Boomerang lens
values, so Boomerang gives synthesized lenses the same first-class status as hand-written ones. To
evaluate the effectiveness of QREs, QRE lenses, and QRE lens synthesis, we
conducted experiments to answer the following questions:
%
\begin{itemize}
  \item {\bf Ease of use.} Does synthesizing QRE lenses from QREs permit an
  easier development process than writing lenses by hand?
  Does synthesizing QRE lenses from QREs permit an easier development
  process than manually writing canonizers and then synthesizing lenses between
  their canonized forms?  
  \item {\bf Performance.} Is the synthesis algorithm/implementation fast enough to be
  used as part of a standard development process?
\end{itemize}
%
All evaluations were performed on a 2.5 GHz Intel Core i7 processor with 16 GB
of 1600 MHz DDR3 running macOS High Sierra.

@@

\BCP{Wrote the rest---comments?}\saz{I swapped the last two paragraphs to end on
  a more exciting note.}
Looking ahead, we are experimenting with larger examples of
quotiented bijective lenses.  So far, the three ``specialized'' QRE
combinators (\kw{squash}, \kw{perm}, and \kw{collapse}) have proved
sufficient, but we expect eventually to encounter examples that will lead us
to consider adding further combinators to supplement these.

Along a different dimension, we would like to investigate whether the techniques
proposed here can be applied to synthesizing quotiented variants of richer
classes of lenses, such as ``classic'' asymmetric lenses~\cite{FosterLenses} or
symmetric lenses~\cite{symmetriclenses}.  The main technical issues are
developing synthesis procedures for the unquotientied variants of a given
class---we are currently attacking the case of symmetric lenses---and showing
that quotiented lenses in this class can be normalized along the lines of
Theorem~\ref{normal form}.
